Setting parameters for biological models with ANIMO

### 2014-02-02 Paper.tex
Adding a workflow graphical representation. Finished the text on the parameter sweep.

% Concludi con esempio di sweep sull'esempio abcd (erk etc): son 3125 simulazioni se lo fai
%  andare solo sui parametri ``slow'', ``medium'', etc e con le 5 interazioni di abcde
We now present an example application of ANIMO's parameter sweep feature on the model shown in Figure~\ref{fig:animo-ui}.
This allows us to make a more exhaustive search than the one presented in Figure~\ref{fig:small-example-feedback},
using an automatic approach instead of a manual search.
A logarithmic parameter sweep was performed, letting the parameter values
vary over the 5 pre-sets ``very slow'' ($k = 0.001$), ``slow'' ($k = 0.002$), \dots, ``very fast'' ($k = 0.016$)
for all 5 interactions in the network. This resulted in $\mbox{\it parameter choices}^{\mbox{\it\scriptsize interactions}} = 5 ^ 5 = 3\,125$ simulations, that were computed
in 43 seconds %43 s with 16 threads. It is 57 with 8 threads (cpu is exploited less than 100%)
on an eight-core Intel\circledR\ Core$^{\mbox{\scriptsize\texttrademark}}$ i7 CPU at 2.80GHz equipped with 4 Gb RAM
and running Ubuntu GNU/Linux 13.10 64bit. The model we consider here is small enough for one simulation to take
less than a second, but the exponential growth of the number of simulations required for a parameter sweep
makes the sheer number of tasks challenging for any processing unit.
% Indeed, a more thorough
% parameter sweep that considers 10 different setting for each of the 5 interactions of the considered
% model would require the computation of $10^5 = 100\,000$ simulations.
Indeed, a more complex model involving 10 interactions instead of 5 would require $5^{10} = 9\,765\,625$ simulations,
which could be computed in approximately a day and a half on an eight-core machine. Yet, such an
estimation would still be better than the twelve days it would take if the simulations were to be computed in sequence.
% the same experiment
% was run executing only one simulation at a time, and it took about 7 minutes and 34 seconds, %453837 ms
% which is in line with the fact that this type of analysis is trivially parallelizable.

@@

@@

% dopo aver presentato gli strumenti che forniamo, definiamo una procedura consigliata per usare ANIMO nel modo più efficace.
% Ovviamente, oltre alle best practices serve anche l'esperienza, sia intesa come conoscenza di biologia che come esperienza
% nel campo della modellazione. Nota che non serve comunque conoscenza di T.A., visto che rimangono sempre nascosti dietro la GUI
% Esempio/case study!
\section{Suggested ANIMO workflow}
In order to better profit from what ANIMO offers to the biologist, we suggest a step-by-step modeling workflow:
\begin{enumerate}
  \item\label{step:def-topology} Define a starting topology, based on literature data and current knowledge on the network.
  \item\label{step:set-rates} Choose the interaction parameters from the pre-set qualitative values,
    based on experience (e.g., protein expression is much slower than phosphorylation).
    The choice on the approximated kinetics can be for the most part based on scenario 1, referring to Section~\ref{sec:parameter-editing}.
  \item\label{step:check-behavior} Check that the network behaves as expected (e.g. addition of EGF makes ERK activity rapidly increase and successively decrease).
  \item Possibly iterate the steps~\ref{step:def-topology} to~\ref{step:check-behavior} until the network behaves as expected.
  \item Compare the model with experimental data.
  \item\label{step:precise-fit} If the model does not fit the data, change the parameters using more precise numerical values and/or scenario settings.
  \item\label{step:param-sweep} If the data fitting is still not satisfactory, apply parameter sweep to (subsets of) the model as described in Section~\ref{sec:parameter-sweep}.
  \item When a good fit is found, add non-determinism to the model and ensure that the
     behavior does not change significantly under reasonably high uncertainty settings.
  \item If the model can fit the data only with very precise parameter settings, or a
    parameter configuration that fits the data cannot be found after a reasonably extensive search
    on the parameter space, change the network topology possibly adding more nodes from literature.
    If no candidates can be found in literature, speculative candidates can be found by ``playing'' with the current
    configuration in ANIMO's user interface.
    We advise to let expert biologists perform this step on their own, or to perform it under their supervision,
    in order to keep the network both simple and realistic. For the same reason, we also advise not to introduce more than a couple of nodes at a time.
  \item Parameters for newly introduced interactions can be chosen as in step~\ref{step:set-rates}, possibly iterating
    steps~\ref{step:def-topology} to~\ref{step:param-sweep} to refine the parameters.
  \item To check how changes in the network are reflected in the behavior of the model, we advise to use the comparison tools
    described in Section~\ref{sec:comparing-versions}.
  \item When a model is obtained that behaves more closely to the experimental data, make it more precisely fitting
    going back to step~\ref{step:precise-fit}.
\end{enumerate}
If the model fits the data reasonably well only after modifications to the initial topology, it may contain some hypothetical or unexpected nodes.
In that case, future experiments can be derived directly from the hypotheses introduced in the model and
help the biologists decide which are the most precise explanations for the experimental observations.

@@

% anche Related work: es. Copasi
% Altro su parameter sweep/sensitivity analysis? Noi usiamo un approccio un po' naif, senza tante pretese, senza tanti numeri.
\section{Conclusions}
ANIMO has been extended with a proper support for parameter synthesis. Now the tool can be used more efficiently
to model biological networks, without concentrating exclusively on their topology.
In particular, the addition of an automatic parameter sweep feature allows the user
to save considerable amounts of time, keeping closer to experimental data without the disadvantages
coming from a hand-made parameter search.

>>>>

% Concludi con esempio di sweep sull'esempio abcd (erk etc): son 3125 simulazioni se lo fai
%  andare solo sui parametri ``slow'', ``medium'', etc e con le 5 interazioni di abcde
We now present an example application of ANIMO's parameter sweep feature on the model shown in Figure~\ref{fig:animo-ui}.
Thanks to parameter sweep, we can make a more exhaustive search than the one presented in Figure~\ref{fig:small-example-feedback},
using an automatic approach instead of a manual trial-and-error process.
A logarithmic parameter sweep was performed, letting the parameter values
vary over the 5 pre-sets ``very slow'' ($k = 0.001$), ``slow'' ($k = 0.002$), ``medium'' ($k = 0.004$), ``fast'' ($k = 0.008$), ``very fast'' ($k = 0.016$)
for all 5 interactions in the network. This resulted in $\mbox{\it parameter choices}^{\mbox{\it\scriptsize interactions}} = 5 ^ 5 = 3\,125$ simulations, that were computed
in 43 seconds %43 s with 16 threads. It is 57 with 8 threads (cpu is exploited less than 100%)
on an eight-core Intel\circledR\ Core$^{\mbox{\scriptsize\texttrademark}}$ i7 CPU at 2.80GHz equipped with 4 Gb RAM
and running Ubuntu GNU/Linux 13.10 64bit. The model we consider here is small enough for one simulation to take
less than a second, but the exponential growth of the number of simulations required for a parameter sweep
can make the sheer number of tasks challenging for any processing unit.
% Indeed, a more thorough
% parameter sweep that considers 10 different setting for each of the 5 interactions of the considered
% model would require the computation of $10^5 = 100\,000$ simulations.
For example, a more complex model involving 10 interactions instead of 5 would require $5^{10} = 9\,765\,625$ simulations,
which could be computed in approximately a day and a half on an eight-core machine.
Still, the possibility to compute simulations in parallel provides a significant help in reducing
the impact of an exponential growth of the space to be explored.
% Yet, such an
% estimation would still be better than the twelve days it would take if the simulations were to be computed in sequence.
% the same experiment
% was run executing only one simulation at a time, and it took about 7 minutes and 34 seconds, %453837 ms
% which is in line with the fact that this type of analysis is trivially parallelizable.

@@

% We note that all scenarios for the 5 interactions were set to the basic scenario 1 for this parameter
% sweep. The graph can be made fit the actual data more precisely by selecting scenario 2 settings for some
% of the interactions. The choice of a scenario has little to no impact on the complexity of the model:
% after choosing scenario 2 for all of the interactions, the whole parameter sweep takes the same time to complete.

@@

% dopo aver presentato gli strumenti che forniamo, definiamo una procedura consigliata per usare ANIMO nel modo più efficace.
% Ovviamente, oltre alle best practices serve anche l'esperienza, sia intesa come conoscenza di biologia che come esperienza
% nel campo della modellazione. Nota che non serve comunque conoscenza di T.A., visto che rimangono sempre nascosti dietro la GUI
% Esempio/case study!
\section{Suggested ANIMO workflow}
In order to better profit from what ANIMO offers to the biologist, we suggest a step-by-step modeling workflow:
\begin{enumerate}
  \item\label{step:def-topology} Define a starting topology, based on literature data and current knowledge on the network.
  \item\label{step:set-rates} Choose the interaction parameters from the pre-set qualitative values,
    based on experience (e.g., protein expression is much slower than phosphorylation).
    The choice on the approximated kinetics can be for the most part based on scenario 1, referring to Section~\ref{sec:parameter-editing}.
  \item\label{step:check-behavior} Check that the network behaves as expected (e.g. addition of EGF makes ERK activity rapidly increase and successively decrease).
  \item Possibly iterate the steps~\ref{step:def-topology} to~\ref{step:check-behavior} until the network behaves as expected.
  \item Compare the model with experimental data.
  \item\label{step:precise-fit} If the model does not fit the data, change the parameters using more precise numerical values and/or scenario settings.
  \item\label{step:param-sweep} If the data fitting is still not satisfactory, apply parameter sweep to (subsets of) the model as described in Section~\ref{sec:parameter-sweep}.
  \item When a good fit is found, add non-determinism to the model and ensure that the
     behavior does not change significantly under reasonably high uncertainty settings.
  \item If the model can fit the data only with very precise parameter settings, or a
    parameter configuration that fits the data cannot be found after a reasonably extensive search
    on the parameter space, change the network topology possibly adding more nodes from literature.
    If no candidates can be found in literature, speculative candidates can be found by ``playing'' with the current
    configuration in ANIMO's user interface.
    We advise to let expert biologists perform this step on their own, or to perform it under their supervision,
    in order to keep the network both simple and realistic. For the same reason, we also advise not to introduce more than a couple of nodes at a time.
  \item Parameters for newly introduced interactions can be chosen as in step~\ref{step:set-rates}, possibly iterating
    steps~\ref{step:def-topology} to~\ref{step:param-sweep} to refine the parameters.
  \item To check how changes in the network are reflected in the behavior of the model, we advise to use the comparison tools
    described in Section~\ref{sec:comparing-versions}.
  \item When a model is obtained that behaves more closely to the experimental data,
  its fitness can be improved going back to step~\ref{step:precise-fit}.
\end{enumerate}
If the model fits the data reasonably well only after modifications to the initial topology, it may contain some hypothetical or unexpected nodes.
In that case, future experiments can be derived directly from the hypotheses introduced in the model and
help the biologists decide which are the most precise explanations for the experimental observations.

@@

% anche Related work: es. Copasi
% Altro su parameter sweep/sensitivity analysis? Noi usiamo un approccio un po' naif, senza tante pretese, senza tanti numeri.
% Possibile future work: vedere meglio altri modi di applicare parameter setting nel contesto dei t.a., usare
% calcolo distribuito per accelerare la ricerca dei parametri
\section{Conclusions}
ANIMO has been extended with a proper support for parameter synthesis. Now the tool can be used more efficiently
to model biological networks, without concentrating exclusively on their topology.
In particular, the addition of an automatic parameter sweep feature allows the user
to save considerable amounts of time, keeping closer to experimental data without the disadvantages
coming from a hand-made parameter search.
